Evaluating the Importance of Initial Conditions for Antarctic Sea Ice Seasonal Predictability with a Fully Coupled Model

### 2025-10-09 analysis/paper/access-ice.tex
Renders

\raggedbottom
\usepackage[section]{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Evaluating Seasonal Predictability of Antarctic Sea Ice and the importance of initial conditions with fully coupled models},
  pdfauthor={Elio Campitelli; Ariaan Purich; Julie Arblaster; Eun-Pa Lim; Matthew Wheeler; Phillip Reid},
  pdfkeywords={sea ice, predictabilty, initial conditions, forecasting},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title[Evaluating Seasonal Predictability of Antarctic Sea Ice and the
importance of initial conditions with fully coupled models]{Evaluating
Seasonal Predictability of Antarctic Sea Ice and the importance of
initial conditions with fully coupled models}

@@

\abstract{Accurate Antarctic sea-ice forecasts are crucial for climate
monitoring and operational planning, yet they remain challenging due to
model biases and complex ice-ocean-atmosphere interactions. We
investigate the role of initial conditions on Antarctic sea-ice
forecasts by comparing hindcasts from two versions of the Australian
Bureau of Meteorology's ACCESS seasonal forecast system: ACCESS-S1 and
ACCESS-S2. They use identical model configuration and differ only in
their initial conditions; primarily in that ACCESS-S2 does not
assimilate sea-ice observations. Our analysis reveals that both systems
experience an extended melt season and delayed growth phase. This leads
to a significant negative sea-ice extent bias which is corrected only in
ACCESS-S1 by the data assimilation system. The impact of the initial
conditions on forecast errors varies dramatically by season: summer and
autumn initial conditions (January-April) provide predictive skill for
up to three months, with February initial conditions being particularly
crucial. In contrast, winter forecasts off the two systems are
statistically indistinguishable after only two weeks. Regional analysis
of forecast skill suggests that this winter predictability barrier is
most dramatic over Eastern Antarctica, where evern ACCESS-S1 shows
negative skill. These findings highlight the critical importance of
comprehensive year-round sampling in predictability studies and suggest
that operational sea-ice data assimilation efforts should prioritize the
summer-autumn period when initial conditions have maximum impact on
forecast skill.}

% keywords
\keywords{sea ice,  predictabilty,  initial conditions,  forecasting}

@@

However, producing accurate Antarctic sea-ice forecasts has been
challenging due to model biases, inherent large variability and
complexity, and they have lagged behind Arctic sea-ice
forecasts\citep{zampieri2019, gao2024a}. Dynamical seasonal forecasts of
summer Antarctic sea ice have been shown to perform worse than
relatively simple statistical methods \citep{massonnet2023} and machine
learning approaches \citep[e.g.][]{dong2024, lin2025}, which also
underscores the need for better understanding of sea-ice dynamics, and
drivers of its variability.

Good initial conditions are generally required for a good forecast,
however, it is not entirely known to what extent accurate sea-ice
initial conditions affect the quality of the forecast and at what
timescales. Exploring seasonal predictions of Arctic sea-ice,
\citet{guemas2016} found that sea-ice initial conditions are important
in autumn to predict summer sea-ice, but the impact wasn't as dramatic
when predicting winter sea-ice. \citet{day2014} also found
seasonally-varying differences in the effect of initialisation, noting
that accurate Arctic sea-ice thickness lead to improved sea-ice
forecasts initialised in July but not when initialised in January.

For the Antarctic, \citet{holland2013} studied the initial-value
predictability of Antarctic sea-ice in a perfect model study using the
CCSM3 model. They found that sea-ice and ocean initial conditions
provide predictive information to forecast sea-ice edge location several
months in advance and that some predictability is retained for up to two
years thanks to ocean heat content anomalies that are advected eastward.
This is in contrast with \citet{marchi2020}, who ran perfect model
experiments to argue that uncertainty in the predicted atmospheric state
and evolution is the main driver of uncertainty in Antarctic sea-ice
extent prediction on seasonal timescales, with sea-ice and ocean initial
conditions having lesser importance. More recently, \citet{morioka2022}
studied decadal forecasts of Antarctic sea-ice and found that
initialising ocean and sea-ice improved the correlation between
simulated and observed sea-ice concentration evolution in the
Amundsen--Bellingshausen Sea. It is hard to compare these studies since
they are based on forecasts initialised at different times of the year
and different frameworks: \citet{holland2013} ran 20 ensemble members
initialised on the 1st of January of a particular year,
\citet{marchi2020} ran forecasts from the 1st of March and 1st of
September and \citet{morioka2022} run forecasts only from the 1st of
March. \citet{marchi2020} also used a coupled ocean--sea-ice model
instead of a fully coupled model like \citet{holland2013} did.
\citet{morioka2022} used observed sea ice initial conditions and
compared with observations, while \citet{marchi2020} and
\citet{holland2013} are perfect model studies.

In October 2021 the Australian Bureau of Meteorology (BoM) upgraded the
Australian Community Climate and Earth System Simulator -- Seasonal
(ACCES-S) from version S1 to S2. While the base model remained the same,
the change in version was focused on using ocean, sea-ice and land
initial conditions generated by the BoM instead of depending on the UK
Met Office. Crucially, compared to ACCESS-S1, ACCESS-S2 does not
assimilate sea-ice observations, so sea ice is only affected by the
ocean and atmospheric data assimilation via the coupled integration.

Since model configuration is identical between ACCESS-S1 and ACCESS-S2,
they form a sort of ``natural experiment'' where the same forecasting
model was run over a long period of time with multiple ensemble
forecasts initialised throughout the year, with the only difference
being the initial conditions. This provides an opportunity to test the
effect of sea-ice initial conditions on the forecast of sea-ice
concentrations and the climate.

In this study we compare sea-ice hindcasts produced by ACCESS-S1 and
ACCESS-S2. We focus on seasonality of errors and biases and the effect
of the data assimilation system. This comparison will also inform future
work with the prediction system as a research tool to better understand
the dynamics and variability of the Antarctic sea ice and its impacts on
the climate system as well as to explore the potential of using its
sea-ice forecasts for decision-making. The work will also serve as a
benchmark for future prediction systems to attempt to improve upon.

@@

Both systems take atmospheric initial conditions derived from
ERA-interim \citep{dee2011} for their hindcasts. The main difference
between the hindcasts of the two systems are the ocean and sea-ice
initial conditions. ACCESS-S1 ocean initial conditions come from the Met
Office FOAM system, which uses a multivariate, incremental
three-dimensional variational (3D-Var), first-guess-at-appropriate-time
(FGAT) data assimilation scheme \citep{waters2015} and assimilates sea
surface temperature (SST), sea surface height (SSH), in situ temperature
and salinity profiles, and sea-ice concentration. ACCESS-S2, on the
other hand, is initialised from ocean conditions generated by the BoM
weakly coupled ensemble data assimilation scheme described in
\citet{wedd2022}. This scheme uses an optimal interpolation method and
assimilates temperature and salinity profiles from EN4 \citep{good2013}.
SSTs are nudged to Reynolds OISSTv2.1 \citep{reynolds2007} in areas
where SSTs are over 0ºC and Sea Surface Salinity is weakly nudged to the
World Ocean Atlas 2013 climatology \citep{zweng2013}.

Of most relevance for this work, sea-ice concentrations are not
assimilated in ACCESS-S2. Assimilation cycles are performed daily. The
coupled model runs for 24 hours initialised from the previous cycle.
Then the restart file fields of the ocean component are used as first
guess in the data assimilation cycle and the innovations are used to
build the next ocean initial conditions for the following cycle. The
atmosphere fields from that daily integration are not used and instead
the model atmosphere is initialised using ERA-Interim. The sea-ice
initial conditions for the next cycle are the unaltered output of the
previous daily integration. Then the cycle starts again and the coupled
model runs for another 24 hours. During this integration the sea-ice
component is affected by the ocean innovations and the new atmosphere
initial conditions via the coupler.

The ACCESS-S2 hindcast set used in this study runs for the period
1981--2018. Each hindcast consists of nine ensemble members built from
three consecutive 3-member forecasts initialised at the first of every
month and the two previous days and run for 279 days. The ACCESS-S1
hindcast set is built by perturbing the atmospheric fields only with a
random field perturbation \citep{hudson2017} and runs for 217 days for
the period 1990--2012. We analise the ensemble mean hindcasts unless
otherwise specified.

Anomalies for each hindcast set are taken with respect to their own
climatology specific to each initialisation date and forecast lead time,
for the period 1990--2012. This serves as a first-order correction of
model bias and drift.

@@

Figure~\ref{fig-hindcast-extent} shows the mean sea-ice extent of the
ACCESS-S2 and ACCESS-S1 hindcasts (row a) and their differences from the
mean sea-ice extent of NOAA/NSIDC CDRV4 (row b). Mean extent at the
first of every month is indicated with circles for the initial
conditions and with triangles for the longest lead time possible for
each model (between 274 and 277 days for ACCESS-S2 and between 213 and
216 days for ACCESS-S1). At this long lead time, the information of the
initial conditions is essentially lost and the forecast reverts to each
model's preferred equilibrium state.

ACCESS-S2 initial conditions (circles; Figure~\ref{fig-hindcast-extent}
column 1) show an overall negative bias, especially in the late
summer-early autumn, while ACCESS-S1 initial conditions (circles in
Figure~\ref{fig-hindcast-extent} column 2) are very close to
observations, as expected from the assimilation of sea-ice observations
to produce the initial conditions of ACCESS-S1. Both systems'
equilibrium states (triangles) show negative biases of sea-ice extent,
particularly in the growth phase of late-autumn and winter months. This
is due primarily to the melt season being longer than in observations
and with faster melt between January and March and the growing seasons
being shorter with slower growth during March and April. This is then
followed by faster growth between May and July
(Figure~\ref{fig-mean-growth}). Many sea-ice models exhibit this
systematic underestimation during the sea-ice minimum and early freezing
season \citep{massonnet2023}, which could indicate problems in the
representation of thermodynamics in the model \citep{zampieri2019}. It
is also not surprising that both forecasting systems converge to a
similar equilibrium state because they share the same model formulation.

The difference between the initial conditions (circles) and the model
equilibrium state (triangles) can be mostly attributed to the effect of
data assimilation, which in ACCESS-S2 is due solely to atmospheric and
oceanic data assimilation. From May to October, in ACCESS-S2 circles are
closer to observations than to the triangles, indicating that the
information from the ocean and atmosphere data assimilation is affecting
sea ice and improving the initial conditions. During these months,
ACCESS-S1 can overestimate the sea-ice extent at short lead time. For
the rest of the year circles are overlaid with triangles in ACCESS-S2,
indicating that the ocean and atmosphere data assimilation is not
affecting sea ice and that this component of the model is virtually
free-running.

@@

\caption{\label{fig-mean-growth}Mean daily sea-ice extent growth
(10\^{}6 km\^{}2/day) in ACCESS-S1 (green) and ACCESS-S2 (purple)
hindcasts and observations (black), computed as the mean daily
differences in sea-ice extent between each date and the next for each
forecast month. Values are smoothed with a 11-day running mean.}

@@

ACCESS-S1 has a comparatively smaller overall bias
(Figure~\ref{fig-bias-2}). The largest values are found between April
and June, when the faster growth resutls in large positive bias along
the sea-ice edge, and in January, when the faster melt leads to large
negative bias in the Weddell and Admunsen seas.

@@

Unexpectedly, for ACCESS-S2, RMSE improves with lead time, even though
the correlation degrades with lead time. This is puzzling behaviour that
goes contrary to what is usually seen in prediction models. The
explanation seems to be the mentioned increased interannual variability.
Figure~\ref{fig-extent-sd} shows monthly mean sea-ice extent of the
forecasts as a function of lead time compared with observations.
ACCESS-S1 standard deviation lies within the observed standard deviation
regardless of lead time, while ACCESS-S2 standard deviation is more than
twice that of observations at zero lead time and only approaches the
observed value at 9 month lead time for most months.

@@

To analyse the spatial distribution of the model error, we computed the
RMSE of zonal mean sea-ice concentration anomalies on 15 slices of
\texttt{diff(unique(rmse\_lon\_mean\$lon)){[}1{]}}° longitude span for
each forecasting system. We control for some areas being naturally
easier to forecast than others by computing the RMSE skill score with
the climatological forecast RMSE as reference.

@@

In Figure 5 the mean error was shown. Figure~\ref{fig-initial-spread}
column 1 shows the mean standard deviation of errors among ensemble
members at various lead times. At one day lead time
(Fig.~\ref{fig-initial-spread} a.1) ACCESS-S2 has a slightly larger
spread than ACCESS-S1 due to the way that ensemble members are
generated. ACCESS-S1 ensemble members are generated by adding random
field perturbations to the atmosphere only, which then are transferred
to the other components via the coupled simulation \citep{hudson2017}.
With this scheme, ensemble members are all but guaranteed to be
underdispersed in the ocean and sea-ice components. The time-lag
ensemble used for ACCESS-S2 ensures greater spread. This difference is
gone after about just two days, and both systems have a comparable
spread in ensemble member error afterwards
(Fig.~\ref{fig-initial-spread} b1 and c1).

@@

Sea-ice forecasts from the ACCESS-S2 system show a significant low
extent bias, particularly during late summer and early autumn. This bias
is attributed to a faster and longer melt season between January and
March, and slower growth between March and April. This underestimation
during the minimum and early freezing season is a common issue in many
seasonal-to-subseasonal (S2S) systems, suggesting potential problems
either with the model's thermodynamic representation or with short wave
radiation forcing, as shown in other climate models
\citep{zampieri2019, roach2020}. Even though ACCESS-S2 shares the same
model formulation with ACCESS-S1, the latter does not suffer from this
bias, indicating that assimilating sea-ice concentrations successfully
corrects for the negative bias that exists in the free-running model.

Ensemble spread grows quickly even when perturbations are only
implemented in the atmosphere component (in ACCESS-S1), indicating that
sea-ice is indeed responding quickly to atmospheric perturbations.
However, our analysis suggests that the atmosphere and ocean data
assimilation implemented in ACCESS-S2 is only effectively influencing
sea-ice initial conditions from June to October, while the rest of the
year, the sea-ice component runs virtually free, reverting to its biased
equilibrium state.

@@

Although ACCESS-S1 only assimilates sea-ice concentration, it is clear
that sea-ice thickness is also affected. ACCESS-S1 simulates
significantly thicker ice than ACCESS-S2 and in both systems sea-ice is
thicker at shorter lead times than at longer lead times. Both the
explicit data assimilation in ACCESS-S2 and the effects of atmospheric
and oceanic data assimilation in ACCESS-S1 might be nudging simulated
sea ice to be thicker than the model equilimbrium state.\\
We believe that the thinner sea-ice in ACCESS-S2 is a partial cause for
the large sea-ice extent variance, but other mechanisms, such as
unbalanced initial conditions might be important.

Given that ACCESS-S2 sea-ice extent is essentially not initialised by
the observations, comparing its forecasts with those of ACCESS-S1 allows
us to estimate the time-scale for which initial conditions are
important. We find that initial conditions affect Antarctic sea-ice
forecasts in ther order of a few months, but that effect is seasonally
dependent. January to April initial conditions improve forecasts for up
to 3 months with, February initial conditions in particular are shown to
be crucial for determining sea-ice evolution at least up to May. Arctic
sea-ice forecasts also show greater sensitivity to initial conditions in
boreal summer, compared with boreal winter \citep{day2014, bunzel2016},
suggesting a similar mechanism might be playing a role.

@@

\caption{\label{fig-mean-growth-osi}Mean daily sea-ice extent growth
(10\^{}6 km\^{}2/day) in ACCESS-S1 (green) and ACCESS-S2 (purple)
hindcasts and observations (black), computed as the mean daily
differences in sea-ice extent between each date and the next for each
forecast month. Values are smoothed with a 11-day running mean.}

>>>>

\raggedbottom
\usepackage[section]{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Evaluating the Importance of Initial Conditions for Antarctic Sea Ice Seasonal Predictability with a Fully Coupled Model},
  pdfauthor={Elio Campitelli; Ariaan Purich; Julie Arblaster; Eun-Pa Lim; Matthew Wheeler; Phillip Reid},
  pdfkeywords={sea ice, predictability, initial
conditions, forecasting},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title[Evaluating the Importance of Initial Conditions for Antarctic Sea
Ice Seasonal Predictability with a Fully Coupled Model]{Evaluating the
Importance of Initial Conditions for Antarctic Sea Ice Seasonal
Predictability with a Fully Coupled Model}

@@

\abstract{Accurate Antarctic sea ice forecasts are crucial for climate
monitoring and operational planning, yet they remain challenging due to
model biases and complex ice-ocean-atmosphere interactions. The two
versions of the Australian Bureau of Meteorology's ACCESS seasonal
forecast system, ACCESS-S1 and ACCESS-S2, use identical model
configuration and differ only in their initial conditions; primarily in
that ACCESS-S2 does not assimilate sea-ice observations. This makes for
a good ``natural experiment'' to analyse the role of initial conditions
on Antarctic sea ice forecasts using a long run and a moderately large
ensemble with a fully coupled model. Our analysis reveals that both
systems experience an extended melt season and delayed growth phase.
This leads to a significant negative sea-ice extent bias which is
corrected only in ACCESS-S1 by the data assimilation system. The impact
of the initial conditions on forecast errors varies dramatically by
season: summer and autumn initial conditions (January-April) provide
predictive skill for up to three months, with February initial
conditions being particularly crucial. In contrast, winter forecasts of
the two systems are statistically indistinguishable after only two
weeks. Regional analysis of forecast skill suggests that this winter
predictability barrier is most dramatic over Eastern Antarctica, where
even ACCESS-S1 shows negative skill. These findings highlight the
critical importance of comprehensive year-round sampling in
predictability studies and suggest that operational sea-ice data
assimilation efforts should prioritize the summer-autumn period when
initial conditions have maximum impact on forecast skill.}

% keywords
\keywords{sea ice,  predictability,  initial conditions,  forecasting}

@@

However, Antarctic sea ice forecasts have lagged behind Arctic sea ice
forecasts due to model biases, inherent large variability and complexity
\citep{zampieri2019, gao2024a}. Dynamical seasonal forecasts of summer
Antarctic sea ice have been shown to perform worse than relatively
simpler statistical methods \citep{massonnet2023} and machine learning
approaches (e.g. \citet{dong2024}, \citet{lin2025}), which also
underscores the need for better understanding of sea-ice dynamics, and
drivers of its variability.

Good initial conditions are generally required for a good forecast,
however, it is not entirely known to what extent accurate sea ice
initial conditions affect the quality of the forecast and at what
timescales. Exploring seasonal predictions of Arctic sea ice,
\citet{guemas2016} found that sea ice initial conditions are important
in autumn to predict summer sea ice, but the impact wasn't as dramatic
when predicting winter sea ice. \citet{day2014} also found
seasonally-varying differences in the effect of initialisation, noting
that accurate Arctic sea-ice thickness lead to improved sea ice
forecasts initialised in July but not when initialised in January.

For the Antarctic, \citet{holland2013} studied the initial-value
predictability of Antarctic sea ice in a perfect model study using the
CCSM3 model. They found that sea-ice and ocean initial conditions
provide predictive information to forecast sea-ice edge location several
months in advance and that some predictability is retained for up to two
years thanks to ocean heat content anomalies that are advected eastward.
This is in contrast with \citet{marchi2020}, who ran perfect model
experiments to argue that uncertainty in the predicted atmospheric state
and evolution is the main driver of uncertainty in Antarctic sea-ice
extent prediction on seasonal timescales, with sea-ice and ocean initial
conditions having lesser importance. More recently, \citet{morioka2022}
studied decadal forecasts of Antarctic sea ice and found that
initialising ocean and sea ice improved the correlation between
simulated and observed sea-ice concentration evolution in the
Amundsen--Bellingshausen Sea. It is hard to compare these studies since
they are based on forecasts initialised at different times of the year
and different frameworks: \citet{holland2013} ran 20 ensemble members
initialised on the 1st of January of a particular year,
\citet{marchi2020} ran forecasts from the 1st of March and 1st of
September and \citet{morioka2022} run forecasts only from the 1st of
March. \citet{marchi2020} also used a coupled ocean--sea-ice model
instead of a fully coupled model like \citet{holland2013} did.
\citet{morioka2022} used observed sea ice initial conditions and
compared with observations, while \citet{marchi2020} and
\citet{holland2013} are perfect model studies.

In October 2021 the Australian Bureau of Meteorology (BoM) upgraded the
Australian Community Climate and Earth System Simulator -- Seasonal
(ACCESS-S) from version S1 to S2. While the base model remained the
same, the change in version was focused on using ocean, sea-ice and land
initial conditions generated by the BoM instead of depending on the UK
Met Office. Crucially, compared to ACCESS-S1, ACCESS-S2 does not
assimilate sea-ice observations, so sea ice is only affected by the
ocean and atmospheric data assimilation via the coupled integration.

Since model configuration is identical between ACCESS-S1 and ACCESS-S2,
they form a sort of ``natural experiment'' where the same forecasting
model was run over a long period of time with multiple ensemble
forecasts initialised throughout the year, with the only difference
being the initial conditions. This provides an opportunity to test the
effect of sea ice initial conditions on the forecast of sea-ice
concentrations and the climate.

In this study we compare sea ice hindcasts produced by ACCESS-S1 and
ACCESS-S2. We focus on seasonality of errors and biases and the effect
of the data assimilation system. This comparison will also inform future
work with the prediction system as a research tool to better understand
the dynamics and variability of the Antarctic sea ice and its impacts on
the climate system as well as to explore the potential of using its sea
ice forecasts for decision-making. The work will also serve as a
benchmark for future prediction systems to attempt to improve upon.

@@

Both systems take atmospheric initial conditions derived from
ERA-interim \citep{dee2011} for their hindcasts. The main difference
between the hindcasts of the two systems are the ocean and sea ice
initial conditions. ACCESS-S1's ocean and sea ice initial conditions
come from the Met Office FOAM system, which uses a multivariate,
incremental three-dimensional variational (3D-Var),
first-guess-at-appropriate-time (FGAT) data assimilation scheme
\citep{waters2015} and assimilates sea surface temperature (SST), sea
surface height (SSH), in situ temperature and salinity profiles, and
satellite observations of sea-ice concentration using the EUMETSAT
OSISAF product described in the next section. ACCESS-S2, on the other
hand, is initialised from ocean conditions generated by the BoM weakly
coupled ensemble data assimilation scheme described in \citet{wedd2022}.
This scheme uses an optimal interpolation method and assimilates
temperature and salinity profiles from EN4 \citep{good2013}. SSTs are
nudged to Reynolds OISSTv2.1 \citep{reynolds2007} in areas where SSTs
are over 0ºC and Sea Surface Salinity is weakly nudged to the World
Ocean Atlas 2013 climatology \citep{zweng2013}.

Of most relevance for this work, sea-ice concentrations are not
assimilated in ACCESS-S2. Assimilation cycles are performed daily. The
coupled model runs for 24 hours initialised from the previous cycle.
Then the restart file fields of the ocean component are used as first
guess in the data assimilation cycle and the innovations are used to
build the next ocean initial conditions for the following cycle. The
atmosphere fields from that daily integration are not used and instead
the model atmosphere is initialised using ERA-Interim. The sea ice
initial conditions for the next cycle are the unaltered output of the
previous daily integration. Then the cycle starts again and the coupled
model runs for another 24 hours. During this integration the sea-ice
component is affected by the ocean innovations and the new atmosphere
initial conditions via the coupler.

The ACCESS-S1 hindcast set is made up of nine members created by
perturbing the atmospheric fields only with a random field perturbation
\citep{hudson2017} and runs for 217 days for the period 1990--2012
initialised at the first of every month. The ACCESS-S2 hindcast set used
in this study runs for the period 1981--2018. Each hindcast consists of
nine ensemble members built from three consecutive 3-member forecasts
created in the same manner as ACCESS-S1 members initialised at the first
of every month and the two previous days and run for 279 days. We
analyse the ensemble mean hindcasts unless otherwise specified.

Anomalies for each hindcast set are taken with respect to their own
climatology specific to each initialisation date and forecast lead time,
for the period 1990--2012. This serves as a first-order correction of
model bias and drift. For monthly means, we define ``0 lead time
months'' as the monthly mean forecast of the same month of
initialisation.

@@

Figure~\ref{fig-hindcast-extent} shows mean sea-ice extent of the
ACCESS-S1 and ACCESS-S2 hindcasts (row a) and their differences from
mean sea-ice extent of NOAA/NSIDC CDRV4 (row b). Mean extent at the
first of every month is indicated with circles for the initial
conditions and with triangles for the longest lead time possible for
each model (between 274 and 277 days for ACCESS-S2 and between 213 and
216 days for ACCESS-S1). At this long lead time, information of the
initial conditions is essentially lost and the forecast reverts to each
model's preferred equilibrium state.

ACCESS-S2 initial conditions (circles in Fig.~\ref{fig-hindcast-extent}
column 2) show an overall negative bias, especially in the late
summer-early autumn, while ACCESS-S1 initial conditions (circles in
Fig.~\ref{fig-hindcast-extent} column 1) are very close to observations,
as expected from the assimilation of sea-ice observations to produce the
initial conditions of ACCESS-S1. Both systems' equilibrium states
(triangles) show negative biases of sea-ice extent, particularly in the
growth phase of late-autumn and winter months. This is due primarily to
the melt season being longer than in observations and with faster melt
between January and March and the growing seasons being shorter with
slower growth during March and April. This is then followed by faster
growth between May and July (Figure~\ref{fig-mean-growth}). Many sea-ice
models exhibit this systematic underestimation during the sea-ice
minimum and early freezing season \citep{massonnet2023}, which could
indicate problems in the representation of thermodynamics in the model
\citep{zampieri2019}. It is also not surprising that both forecasting
systems converge to a similar equilibrium state because they share the
same model formulation.

The difference between the initial conditions (circles) and the model
equilibrium state (triangles) can be mostly attributed to the effect of
data assimilation, which in ACCESS-S2 is due solely to the coupling of
sea-ice with the atmosphere and the ocean. From May to October, in
ACCESS-S2 circles are closer to observations than the triangles are,
indicating that the information from the ocean and atmosphere data
assimilation is affecting sea ice and improving the initial conditions.
During these months, ACCESS-S1 can overestimate the sea-ice extent at
short lead time. For the rest of the year circles are overlaid with
triangles in ACCESS-S2, indicating that the ocean and atmosphere data
assimilation is not affecting sea ice and that this component of the
model is virtually free-running.

@@

\caption{\label{fig-mean-growth}Mean daily sea-ice extent growth
(\(10^6 km^2/day\)) in ACCESS-S1 (green) and ACCESS-S2 (purple)
hindcasts and observations (black), computed as the mean daily
differences in sea-ice extent between each date and the next for each
forecast month. Values are smoothed with a 11-day running mean.}

@@

ACCESS-S1 has a comparatively smaller overall bias
(Figure~\ref{fig-bias-2}). The largest values are found between April
and June, when the faster growth results in large positive bias along
the sea-ice edge, and in January, when the faster melt leads to large
negative bias in the Weddell and Admunsen seas.

@@

Unexpectedly, for ACCESS-S2, RMSE improves with lead time, even though
the correlation degrades with lead time. This is puzzling behaviour that
goes contrary to what is usually seen in prediction models. The
explanation seems to be the mentioned increased interannual variability.
Figure~\ref{fig-extent-sd} shows the interannual standard deviation of
monthly sea-ice extent of the forecasts as a function of lead time
compared with observations. ACCESS-S1 standard deviation lies within the
observed standard deviation regardless of lead time, while ACCESS-S2
standard deviation is more than twice that of observations at zero lead
time and only approaches the observed value at 9 month lead time for
most months.

@@

To analyse the spatial distribution of the model error, we computed the
RMSE of zonal mean sea-ice concentration anomalies on 15 slices of 24°
longitude span for each forecasting system. We control for some areas
being naturally easier to forecast than others by computing the RMSE
skill score with the climatological forecast RMSE as reference.

@@

In Figure~\ref{fig-rmse} the mean error was shown.
Figure~\ref{fig-initial-spread} column 1 shows the mean standard
deviation of errors among ensemble members at various lead times. At one
day lead time (Fig.~\ref{fig-initial-spread} a.1) ACCESS-S2 has a
slightly larger spread than ACCESS-S1 due to the way that ensemble
members are generated. ACCESS-S1 ensemble members are generated by
adding random field perturbations to the atmosphere only, which then are
transferred to the other components via the coupled simulation
\citep{hudson2017}. With this scheme, ensemble members are all but
guaranteed to be underdispersed in the ocean and sea-ice components. The
time-lag ensemble used for ACCESS-S2 ensures greater spread. This
difference is gone after about just two days, and both systems have a
comparable spread in ensemble member error afterwards
(Fig.~\ref{fig-initial-spread} b1 and c1).

@@

Sea ice forecasts from the ACCESS-S2 system show a significant low
extent bias, particularly during late summer and early autumn. This bias
is attributed to a faster and longer melt season between January and
March, and slower growth between March and April. This underestimation
during the minimum and early freezing season is a common issue in many
seasonal-to-subseasonal (S2S) systems, suggesting potential problems
either with the model's thermodynamic representation or with short wave
radiation forcing, as shown in other climate models
\citep{zampieri2019, roach2020}. Even though ACCESS-S2 shares the same
model formulation with ACCESS-S1, the latter does not suffer from this
bias, indicating that assimilating sea-ice concentrations successfully
corrects for the negative bias that exists in the free-running model.

Ensemble spread grows quickly even when perturbations are only
implemented in the atmosphere component (in ACCESS-S1), indicating that
sea-ice is indeed responding quickly to atmospheric perturbations.
However, our analysis suggests that the atmosphere and ocean data
assimilation implemented in ACCESS-S2 is only effectively influencing
sea ice initial conditions from June to October, while the rest of the
year, the sea-ice component runs virtually free, reverting to its biased
equilibrium state.

@@

Although ACCESS-S1 only assimilates sea-ice concentration, it is clear
that sea-ice thickness is also affected. ACCESS-S1 simulates
significantly thicker ice than ACCESS-S2 and in both systems sea-ice is
thicker at shorter lead times than at longer lead times. Both the
explicit data assimilation in ACCESS-S2 and the effects of atmospheric
and oceanic data assimilation in ACCESS-S1 might be nudging simulated
sea ice to be thicker than the model equilimbrium state.\\
We believe that the thinner sea ice in ACCESS-S2 is a partial cause for
the large sea-ice extent variance, but other mechanisms, such as
unbalanced initial conditions might be important.

Given that ACCESS-S2 sea-ice extent is essentially not initialised by
the observations, comparing its forecasts with those of ACCESS-S1 allows
us to estimate the time-scale for which initial conditions are
important. We find that initial conditions affect Antarctic sea ice
forecasts in the order of a few months, but that effect is seasonally
dependent. January to April initial conditions improve forecasts for up
to 3 months with, February initial conditions in particular are shown to
be crucial for determining sea-ice evolution at least up to May. Arctic
sea ice forecasts also show greater sensitivity to initial conditions in
boreal summer, compared with boreal winter \citep{day2014, bunzel2016},
suggesting a similar mechanism might be playing a role.

@@

\caption{\label{fig-mean-growth-osi}Mean daily sea-ice extent growth
(\(10^6 km^2/day\)) in ACCESS-S1 (green) and ACCESS-S2 (purple)
hindcasts and observations (black), computed as the mean daily
differences in sea-ice extent between each date and the next for each
forecast month. Values are smoothed with a 11-day running mean.}
